Fluent statistical computing interfaces for biological data analysis

### 2020-03-24 Rmd/01-intro.Rmd
typos

Exploratory data analysis (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all it’s messes and 
uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
It is an iterative process involving computation and visualization,
leading to new hypotheses that can be tested and formalised using 
statistical modelling. As datasets grow in complexity and become
increasingly heterogeneous and multidimensional, 
the use of EDA becomes vital to ensure the integrity and quality of analysis 
outputs. This is certainly true in high-throughput biological 
data sceince, where constraints on computation time and memory, in addition
to the analyst's time, results in EDA becoming increasingly difficult and 
neglected.

@@

- talking about chapters 3 and 4 here
- having the grammar doesn't impair interoperability, in fact we show it
can be used to glue two different approches
- the long form representation can be taken even further... 
- recasting EDA problems as search, we want to find regions measured along the
genome that display a characteristic of interest, and then link that to raw 
data plots.

>>>>

Exploratory data analysis (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all it’s messes and 
uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
It is an iterative process involving computation and visualization,
leading to new hypotheses that can be tested and formalised using 
statistical modelling. As datasets grow in complexity and become
increasingly heterogeneous and multidimensional, 
the use of EDA becomes vital to ensure the integrity and quality of analysis 
outputs. This is certainly true in high-throughput biological 
data science, where constraints on computation time and memory, in addition
to the analyst's time, results in EDA becoming increasingly difficult and 
neglected.

@@

- talking about chapters 3 and 4 here
- having the grammar doesn't impair interoperability, in fact we show it
can be used to glue two different approaches
- the long form representation can be taken even further... 
- recasting EDA problems as search, we want to find regions measured along the
genome that display a characteristic of interest, and then link that to raw 
data plots.

### 2020-03-24 Rmd/05-embeddings.Rmd
typos

We have combined the tour view in a side by side layout with 
a scatter plot view that represents the output of an embedding algorithm. 
These views are linked; analysts can brush regions or
highlight collections of points in either view. Linked highlighting can
be performed when points have been previously labeled according to some
discrete structure, i.e. cluster labels are available. This is achieved
via the analyst clicking on groups in the legend, which causes unclicked
groupings to have their points become less opaque. Consequently, 
simple linked highlighting can alleviate a known downfall of methods such as 
UMAP or t-SNE: that is distances between clusters are misleading. By
highlighting corresponding clusters in the tour view, the analyst can
see the relationship between clusters, and therefore obtain a more accurate 
representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming
the embedding is representative of the high-dimensional dataset it was
computed from. Defining what it means for embedding to be 'representative` or
'faithful' to high-dimensional data is ill-posed and depends on the underlying
taks an analyst is trying to achieve. At the very minimum, we are interested
in distortions and diffusions of the high-dimensional data. Distortions
occur when points that are near each other in the embedding view are 
far from each other in the original dataset. This implies that the embedding 
is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
is reliant on the distance metric used; distortions and diffusions can
be thought of as the preservation of distances or the nearest neighbors
graphs between the high-dimensional space and the embedding space. As distances
can be noisy in high-dimensions, ranks can be used instead as has been 
proposed by ... Identifying distortions and diffusions allows an analyst
to investigate the quality of their embedding and revise them iteratively. We
propose this can be done visually using our side-by-side tour and embedding
views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
brushes and brush compostions.

>>>>

We have combined the tour view in a side by side layout with 
a scatter plot view that represents the output of an embedding algorithm. 
These views are linked; analysts can brush regions or
highlight collections of points in either view. Linked highlighting can
be performed when points have been previously labeled according to some
discrete structure, i.e. cluster labels are available. This is achieved
via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
simple linked highlighting can alleviate a known downfall of methods such as 
UMAP or t-SNE: that is distances between clusters are misleading. By
highlighting corresponding clusters in the tour view, the analyst can
see the relationship between clusters, and therefore obtain a more accurate 
representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming
the embedding is representative of the high-dimensional dataset it was
computed from. Defining what it means for embedding to be 'representative` or
'faithful' to high-dimensional data is ill-posed and depends on the underlying
task an analyst is trying to achieve. At the very minimum, we are interested
in distortions and diffusions of the high-dimensional data. Distortions
occur when points that are near each other in the embedding view are 
far from each other in the original dataset. This implies that the embedding 
is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
is reliant on the distance metric used; distortions and diffusions can
be thought of as the preservation of distances or the nearest neighbors
graphs between the high-dimensional space and the embedding space. As distances
can be noisy in high-dimensions, ranks can be used instead as has been 
proposed by ... Identifying distortions and diffusions allows an analyst
to investigate the quality of their embedding and revise them iteratively. We
propose this can be done visually using our side-by-side tour and embedding
views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
brushes and brush composition.
